---
title: "Data Science for Public Policy (assignment04)"
author: "Sejin Kim (Net ID: sk2723)"
format: html
editor: source
embed-resources: true
execute:
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

## 1. Educational Attainment Across U.S. States

This visualization compares the share of adults age 25 and older with a bachelor's degree or higher across the 50 U.S. states in 2024. Educational attainment varies considerably across states, providing an overview of geographic differences in higher education attainment.  
(Data source: [U.S. Census Bureau, 2024 ACS 1-Year Estimates, Table B15003](https://data.census.gov/table?q=B15003&g=010XX00US$0400000&y=2024))

```{r}
#| fig-height: 10

library(tidyverse)

education <- read_csv("data/B15003.csv")
glimpse(education)

education <- education |>
  select(`Label (Grouping)`, ends_with("Estimate")) |>
  rename(
    label = `Label (Grouping)`
  ) |>
  rename_with(
    ~ str_remove(.x, "!!Estimate$"),
    ends_with("!!Estimate")
  ) |>
  pivot_longer(
    cols = -label,
    names_to = "state",
    values_to = "population"
  ) |>
  mutate(label = str_trim(label))


education_bachelor <- education |>
  filter(
    label %in% c(
      "Total:",
      "Bachelor's degree",
      "Master's degree",
      "Professional school degree",
      "Doctorate degree"
    )
  ) |>
  group_by(state) |>
  summarize(
    total = population[label == "Total:"],
    bachelor_plus = sum(population[label != "Total:"]),
    bachelor_pct = bachelor_plus / total * 100
  ) |>
  filter(!state %in% c("Puerto Rico", "District of Columbia"))

education_bachelor |>
  ggplot(aes(
    x = bachelor_pct,
    y = reorder(state, bachelor_pct)
  )) +
  geom_col(
    fill = "steelblue",
    width = 0.8
  ) +
  scale_x_continuous(
    labels = scales::label_percent(scale = 1)
  ) +
  labs(
    title = "Educational Attainment Across U.S. States",
    subtitle = "Share of adults age 25 and older with a bachelor's degree or higher, 2024",
    x = "Bachelor's degree or higher",
    y = NULL,
    caption = "Source: U.S. Census Bureau, 2024 American Community Survey 1-Year Estimates"
  ) +
  theme_minimal() +
  theme(
    plot.caption = element_text(hjust = 0)
  )
```

## 2. Educational Attainment and Household Income

The first visualization showed substantial differences in educational attainment across states. This visualization examines whether those differences are associated with household income. Each point represents a state, with colors and shapes indicating Census regions.  
(Data source: [U.S. Census Bureau, 2024 ACS 1-Year Estimates, Table B15003](https://data.census.gov/table?q=B19013&g=010XX00US$0400000&y=2024)

```{r}
income <- read_csv("data/B19013.csv")
glimpse(income)

income <- income |>
  select(ends_with("Estimate")) |>
  rename_with(
    ~ str_remove(.x, "!!Estimate$"),
    ends_with("!!Estimate")
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "state",
    values_to = "median_income"
  ) |>
  filter(!state %in% c("District of Columbia", "Puerto Rico"))

education_income <- education_bachelor |>
  left_join(income, by = "state") |>
  mutate(region = case_when(
    state %in% c(
      "Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania") ~ "Northeast",
    state %in% c(
      "Indiana", "Illinois", "Michigan", "Ohio", "Wisconsin", "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota") ~ "Midwest",
    state %in% c(
      "Delaware", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "West Virginia", "Alabama", "Kentucky", "Mississippi", "Tennessee", "Arkansas", "Louisiana", "Oklahoma", "Texas") ~ "South",
    state %in% c(
      "Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming", "Alaska", "California", "Hawaii", "Oregon", "Washington"
    ) ~ "West"
  ))

education_income |>
  ggplot(aes(
    x = bachelor_pct,
    y = median_income
  )) +
  geom_point(
    aes(color = region,
        shape = region),
    size = 3,
    alpha = 0.7,
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black"
  ) +
  scale_shape_manual(
    values = c(
      "Midwest" = 16,
      "Northeast" = 17,
      "South" = 15,
      "West" = 18
    )
  ) +
  scale_x_continuous(
    labels = scales::label_percent(scale = 1)
  ) +
  scale_y_continuous(
    labels = scales::label_dollar()
  ) +
  labs(
    title = "Educational Attainment and Household Income Across the U.S. States",
    subtitle = "Median household income and bachelor's degree attainment by Census region, 2024",
    x = "Bacelor's degree or higher",
    y = "Median household income",
    color = "Census region",
    shape = "Census region",
    caption = "Source: U.S. Census Bureau, 2024 American Community Survey 1-Year Estimates"
  ) +
  theme_minimal() +
  theme(
    plot.caption = element_text(hjust = 0)
  )

```

## 3. 




