---
title: "Data Science for Public Policy (assignment04)"
author: "Sejin Kim (Net ID: sk2723)"
format: html
editor: source
embed-resources: true
execute:
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---
# Education, Gender, and Economic Outcomes in the United States

## 1. Educational Attainment Across U.S. States

This visualization compares the share of adults age 25 and older with a bachelor's degree or higher across the 50 U.S. states in 2024. Educational attainment varies considerably across states, providing an overview of geographic differences in higher education attainment.  
(Data source: [U.S. Census Bureau, 2024 ACS 1-Year Estimates, Table B15003](https://data.census.gov/table?q=B15003&g=010XX00US$0400000&y=2024))

```{r}
#| fig-height: 10

library(tidyverse)

education <- read_csv("data/B15003.csv")
glimpse(education)

education <- education |>
  select(`Label (Grouping)`, ends_with("Estimate")) |>
  rename(
    label = `Label (Grouping)`
  ) |>
  rename_with(
    ~ str_remove(.x, "!!Estimate$"),
    ends_with("!!Estimate")
  ) |>
  pivot_longer(
    cols = -label,
    names_to = "state",
    values_to = "population"
  ) |>
  mutate(label = str_trim(label))


education_bachelor <- education |>
  filter(
    label %in% c(
      "Total:",
      "Bachelor's degree",
      "Master's degree",
      "Professional school degree",
      "Doctorate degree"
    )
  ) |>
  group_by(state) |>
  summarize(
    total = population[label == "Total:"],
    bachelor_plus = sum(population[label != "Total:"]),
    bachelor_pct = bachelor_plus / total * 100
  ) |>
  filter(!state %in% c("Puerto Rico", "District of Columbia"))

education_bachelor |>
  ggplot(aes(
    x = bachelor_pct,
    y = reorder(state, bachelor_pct)
  )) +
  geom_col(
    fill = "steelblue",
    width = 0.8
  ) +
  scale_x_continuous(
    labels = scales::label_percent(scale = 1)
  ) +
  labs(
    title = "Educational Attainment Across U.S. States",
    subtitle = "Share of adults age 25 and older with a bachelor's degree or higher, 2024",
    x = "Bachelor's degree or higher",
    y = NULL,
    caption = "Source: U.S. Census Bureau, 2024 American Community Survey 1-Year Estimates"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(hjust = 0)
  )
```

## 2. Educational Attainment and Household Income

The first visualization showed substantial differences in educational attainment across states. This visualization examines whether those differences are associated with household income. Each point represents a state, with colors and shapes indicating Census regions.  
(Data source: [U.S. Census Bureau, 2024 ACS 1-Year Estimates, Table B15003](https://data.census.gov/table?q=B19013&g=010XX00US$0400000&y=2024)

```{r}
income <- read_csv("data/B19013.csv")
glimpse(income)

income <- income |>
  select(ends_with("Estimate")) |>
  rename_with(
    ~ str_remove(.x, "!!Estimate$"),
    ends_with("!!Estimate")
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "state",
    values_to = "median_income"
  ) |>
  filter(!state %in% c("District of Columbia", "Puerto Rico"))

education_income <- education_bachelor |>
  left_join(income, by = "state") |>
  mutate(region = case_when(
    state %in% c(
      "Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania") ~ "Northeast",
    state %in% c(
      "Indiana", "Illinois", "Michigan", "Ohio", "Wisconsin", "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota") ~ "Midwest",
    state %in% c(
      "Delaware", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "West Virginia", "Alabama", "Kentucky", "Mississippi", "Tennessee", "Arkansas", "Louisiana", "Oklahoma", "Texas") ~ "South",
    state %in% c(
      "Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming", "Alaska", "California", "Hawaii", "Oregon", "Washington"
    ) ~ "West"
  ))

education_income |>
  ggplot(aes(
    x = bachelor_pct,
    y = median_income
  )) +
  geom_point(
    aes(color = region,
        shape = region),
    size = 3,
    alpha = 0.7,
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black"
  ) +
  scale_shape_manual(
    values = c(
      "Midwest" = 16,
      "Northeast" = 17,
      "South" = 15,
      "West" = 18
    )
  ) +
  scale_x_continuous(
    labels = scales::label_percent(scale = 1)
  ) +
  scale_y_continuous(
    labels = scales::label_dollar()
  ) +
  labs(
    title = "Educational Attainment and Household Income Across U.S. States",
    subtitle = "Median household income and bachelor's degree attainment by Census region, 2024",
    x = "Bacelor's degree or higher",
    y = "Median household income",
    color = "Census region",
    shape = "Census region",
    caption = "Source: U.S. Census Bureau, 2024 American Community Survey 1-Year Estimates"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    plot.caption = element_text(hjust = 0)
  )

```

## 3. Gender Differences in Educational Attainment

The first two visualizations examined educational attainment across states and its relationship with household income. This visualization adds a demographic dimension by comparing the share of men and women age 25 and older with a bachelor's degree or higher across states. The distance between the two points represents the gender gap in educational attainment within each state.  
(Data source: [U.S. Census Bureau, 2024 ACS 1-Year Estimates, Table B15002](https://data.census.gov/table?q=B15002&g=010XX00US$0400000&y=2024)

```{r}
#| fig-height: 10

gender <- read_csv("data/B15002.csv")
glimpse(gender)

gender |>
  select(`Label (Grouping)`) |>
  print(n = Inf)

gender <- gender |>
  mutate(
    label = str_trim(`Label (Grouping)`),
    sex = case_when(
      label == "Male:" ~ "Male",
      label == "Female:" ~ "Female"
    )
  ) |>
  fill(sex) |>
  select(-`Label (Grouping)`)

gender <- gender |>
  select(label, sex, ends_with("Estimate")) |>
  rename_with(
    ~ str_remove(.x, "!!Estimate$"),
    ends_with("!!Estimate")
  ) |>
  pivot_longer(
    cols = -c(label, sex),
    names_to = "state",
    values_to = "population"
  )
  
gender_bachelor <- gender |>
  filter(
    label %in% c(
      "Male:",
      "Female:",
      "Bachelor's degree",
      "Master's degree",
      "Professional school degree",
      "Doctorate degree"
    )) |>
  group_by(state, sex) |>
  summarize(
    total = population[label %in% c("Male:", "Female:")],
    bachelor_plus = sum(
      population[label %in% c(
        "Bachelor's degree",
        "Master's degree",
        "Professional school degree",
        "Doctorate degree"
      )]
    ),
    bachelor_pct = bachelor_plus / total * 100) |>
  ungroup() |>
  filter(!state %in% c("District of Columbia", "Puerto Rico"))
  
gender_plot <- gender_bachelor |>
  select(state, sex, bachelor_pct) |>
  pivot_wider(
    names_from = sex,
    values_from = bachelor_pct
  ) |>
  mutate(
    state = reorder(state, Female)
  )

gender_bachelor <- gender_bachelor |>
  mutate(state = factor(state, levels = levels(gender_plot$state)))

gender_plot |>
  ggplot(aes(y = state)) +
  geom_segment(
    aes(
      x = Male,
      xend = Female,
      yend = state
    ),
    linewidth = 0.7,
    color = "grey"
  ) +
  geom_point(
    data = gender_bachelor,
    aes(
      x = bachelor_pct,
      fill = sex
      ),
    shape = 21,
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "Male" = "steelblue",
      "Female" = "indianred"),
      breaks = c("Male", "Female")
    ) +
  scale_x_continuous(
    labels = scales::label_percent(scale = 1)
  ) +
  labs(
    title = "Gender Differences in Educational Attainment Across U.S. States",
    subtitle = "Share of adults age 25 and older with a bachelor's degree or higher, 2024",
    x = "Bachelor's degree or higher",
    y = NULL,
    fill = "Sex",
    caption = "Source: U.S. Census Bureau, 2024 American Community Survey 1-Year Estimates"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(hjust = 0)
  )
```

## 4. Earnings Rise with Education, but the Gender Gap Persists

Median earnings increase with educational attainment for both men and women. However, a gender earnings gap persists across all education levels, with the difference becoming more pronounced among those with bachelor’s and graduate degrees.  
(Data source: [U.S. Census Bureau, 2024 ACS 1-Year Estimates, Table B20004](https://data.census.gov/table?q=B20004&g=010XX00US&y=2024)

```{r}
earning <- read_csv("data/B20004.csv")
glimpse(earning)

earning <- earning |>
  rename(
    label = `Label (Grouping)`,
    median_earning = `United States!!Estimate`,
    moe = `United States!!Margin of Error`
  ) |>
  mutate(
    label = str_trim(label),
    sex = case_when(
      label == "Male:" ~ "Male",
      label == "Female:" ~ "Female"
    )) |>
  fill(sex) |>
  filter(!is.na(sex)) |>
  filter(!label %in% c("Male:", "Female:"))


earning <- earning |>
  mutate(
    moe = parse_number(moe),
    label = factor(
      label,
      levels = c(
        "Less than high school graduate",
        "High school graduate (includes equivalency)",
        "Some college or associate's degree",
        "Bachelor's degree",
        "Graduate or professional degree"
      ),
      labels = c(
        "< High School",
        "High School",
        "Some College /\nAssociate's",
        "Bachelor's",
        "Graduate"
      )))

ggplot(
  earning,
  aes(
    x = label,
    y = median_earning,
    color = sex,
    group = sex
  )
) +
  geom_line(
    linewidth = 0.8
  ) +
  geom_point(
    size = 3,
    shape = 15
  ) +
  scale_y_continuous(
    labels = scales::label_dollar(),
    breaks = seq(20000, 120000, 20000),
    limits = c(20000, 120000)
  ) +
  labs(
    title = "Earnings Rise with Education, but the Gender Gap Persists",
    subtitle = "Median earnings by sex and educational attainment, United States, 2024",
    x = "Educational Attainment",
    y = "Median Earning",
    color = NULL,
    caption = "Source: U.S. Census Bureau, 2024 American Community Survey (ACS)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(hjust = 0))
```




